Insecticide resistance evolution with mixtures and sequences : a model-based explanation.

### 2018-01-02 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
starting to address reviewers comments

## Methods
The simulation represents a population of randomly mixing individuals using standard population genetic approaches to avoid the need to follow every individual. One gene or locus is represented for each insecticide. Each individual has two copies of the gene and there are two potential alleles per locus. Each allele confers either resistance or susceptibility to the insecticide. Thus individuals can either have both resistance genes (RR homozygous), both susceptible genes (SS) or one of each (SR, heterozygous). The combination of the two genes is termed the genotype. Fitness is the main currency of the model, representing how much each genotype survives and reproduces. The fitness of the SS genotype in absence of the insecticide is used as the reference genotype and assigned a value of 1. All other fitnesses are assumed to be less than 1, determined by alleles, exposure to the insecticide and other inputs that can be set in the model. These inputs include the effectiveness of the insecticide against SS genotypes, the dominance of the resistance allele and the ability of resistance to restore fitness of the RR genotype in the presence of the insecticide (see Table 1).

@@

The simulation proceeds through non-overlapping generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness, the latter depending on their absolute fitness (Fig 2) and level of exposure. This acts to make the fitter alleles more common over time. Separate sexes and standard sexual reproduction with recombination is included. There is also the potential to set exposure rates to be different for males and females to represent their different behaviours but we do not explore that parameter here.

@@

1. single insecticide
1. two insecticides used in sequence, replacing the first with a second once the frequency of resistance alleles reaches a threshold of 50%
1. two insecticides used in a mixture (concurrently)

A major aim of this paper is to us modern techniques to make the modelling accessible to a wider community. The model is implemented in R [@RCoreTeam2016], the code is open-source and hosted on Github [@South2017] including the code to generate both the figures and text of this paper. We provide an on-line user interface [@South2017a] (Fig 3) accessing the same code and enabling readers with no coding experience to change inputs and run the model themselves.

@@

### Single insecticide
For single insecticide use, higher values of insecticide effectiveness, exposure, resistance restoration and its dominance all resulted in faster resistance spread (Fig 4 A-D) and thus fewer generations to reach a resistance threshold of 0.5. This makes intuitive sense as all increase the impact of the insecticide and thus increase selection for resistance. Similarly, higher values of starting resistance frequency also resulted in shorter times to resistance thresholds (Fig 5A) through the even simpler mechanism that a smaller change in frequency is required to reach the threshold.

@@

In summary increasing effectiveness can favour mixtures moving from column 1 to 2 in Figure 6, and increasing exposure can favour sequences moving from row 1 to 2 in Figure 6. Reduced male exposure with respect to female, as might be expected for male mosquitoes not seeking blood meals, had a similar effect to changing overall exposure. Reducing male exposure (not shown) increased times-to-resistance and favoured mixtures over sequences.

@@

#### Linkage disequilibrium
It has previously been suggested [@Curtis1985][@WHO2012] that using an insecticide to which there was existing resistance in a mixture with a new insecticide with little resistance could lead to more rapid evolution of resistance to the latter through a process termed linkage disequilibrium. Figure 9 C&D show how our model predicts that linkage disequilibrium does not lead to more rapid evolution of resistance to the newer insecticide. In Figures 9 C&D the newer insecticide with lower starting resistance (in red) leads to faster resistance when used alone (red dotted) than when used in a mixture (red solid).

@@

There are other model behaviours that can be changed by modifying model inputs within the computer code. For example, here we assume that insecticide interaction in a mixture is multiplicative, e.g. if the probability of surviving exposure to insecticide A alone is 0.3 and of surviving insecticide B alone is 0.2, then the probability of surviving exposure to a mixture of A and B is 0.3*0.2 = 0.06. This is consistent with recent work on aphids [@Taillebois2016], although that did also indicate some synergistic effects. Modifying model inputs would allow us to incorporate such synergistic or antagonistic effects of mixtures, i.e. that insecticides used together perform better or worse than when used separately.

>>>>

## Methods
The simulation represents a population of randomly mixing individuals using standard population genetic approaches to avoid the need to follow every individual. One gene or locus is represented for each insecticide. Each individual has two copies of the gene and there are two potential alleles per locus. Each allele confers either resistance or susceptibility to the insecticide. Thus individuals can either have both resistance genes (RR homozygous), both susceptible genes (SS) or one of each (SR, heterozygous). The combination of the two genes is termed the genotype. Fitness is the main currency of the model, representing how much each genotype survives and reproduces. The fitness of the SS genotype in the absence of the insecticide is used as the reference genotype and assigned a value of 1. All other fitnesses are assumed to be less than 1, determined by alleles, exposure to the insecticide and other inputs that can be set in the model. These inputs include the effectiveness of the insecticide against SS genotypes, the dominance of the resistance allele and the ability of resistance to restore fitness of the RR genotype in the presence of the insecticide (see Table 1).

@@

The simulation proceeds through non-overlapping generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness, the latter depending on their absolute fitness (Fig 2) and level of exposure. This acts to make the fitter alleles more common over time. Separate sexes and standard sexual reproduction with recombination is included. There is also the potential to set exposure rates to be different for males and females to represent their different behaviours. Females are more likely to come into contact with insecticides on walls and nets when seeking to feed on humans compared to males that do not. We only explore that input briefly here because a previous analysis indicated it was not a major factor determining differences between insecticide-use strategies [@Levick2017].

@@

1. single insecticide
2. two insecticides used in sequence, replacing the first with a second once the frequency of resistance alleles reaches a threshold of 50%
3. two insecticides used in a mixture (concurrently)

A major aim of this paper is to use modern techniques to make the modelling accessible to a wider community. The model is implemented in R [@RCoreTeam2017], the code is open-source and hosted on Github [@South2017] including the code to generate both the figures and text of this paper relying on the packages knitr[@Xie2017], shiny[@Chang2017] and ggplot2[@Wickham2009]. We provide an on-line user interface [@South2017a] (Fig 3) accessing the same code and enabling readers with no coding experience to change inputs and run the model themselves.

@@

### Single insecticide
For single insecticide use, higher values of insecticide effectiveness, exposure, resistance restoration and dominance of restoration all resulted in faster resistance spread (Fig 4 A-D) and thus fewer generations to reach a resistance threshold of 0.5. This makes intuitive sense as all increase the impact of the insecticide and thus increase selection for resistance. Similarly, higher values of starting resistance frequency also resulted in shorter times to resistance thresholds (Fig 5A) through the even simpler mechanism that a smaller change in frequency is required to reach the threshold.

@@

In summary increasing effectiveness can favour mixtures moving from column 1 to 2 in Figure 6, and increasing exposure can favour sequences moving from row 1 to 2 in Figure 6. We do not show the results here, but reduced male exposure with respect to female, as would be expected for male mosquitoes not seeking blood meals, had a similar effect to reducing overall exposure.

@@

#### Linkage disequilibrium
Positive linkage disequilibrium is said to exist when the frequency of two alleles at different loci is higher than expected if they were just selected individually. It has previously been suggested [@Curtis1985][@WHO2012] that using an insecticide to which there was existing resistance in a mixture with a new insecticide with little resistance could lead to more rapid evolution of resistance to the latter through linkage disequilibrium. We explicitly look at the example of a mixture of a new insecticide and one with pre-existing resistance (Figures 9 C&D). In both cases resistance to the newer insecticide with lower starting resistance (in red) increases faster when it is used alone (red dotted) than when it is used in a mixture (red solid). This suggests that linkage disequilibrium does not restrict the potential benefits of mixtures in all situations. These results support the statement in GPIRM [1] that : "Even though the risk for linkage disequilibrium exists, mixtures may still be the most attractive IRM tool …". The dynamics of LD is too complex to discuss in detail here but interested readers can find more discussion in [@Curtis1985][@WHO2012][@Levick2017].

@@

There are other model behaviours that can be changed by modifying model inputs within the computer code. For example, here we assume that insecticide interaction in a mixture is multiplicative, e.g. if the probability of surviving exposure to insecticide A alone is 0.3 and of surviving insecticide B alone is 0.2, then the probability of surviving exposure to a mixture of A and B is 0.3*0.2 = 0.06. This is consistent with recent work on aphids [@Taillebois2016], although that did also indicate some synergistic effects. Modifying model inputs would allow us to incorporate such synergistic or antagonistic effects of mixtures, i.e. that insecticides used together perform better or worse than when used separately. This would allow the model to represent cross-resistance where a resistance gene provides resistance to more than one insecticide. Similarly behavioural resistance is likely to offer resistance against more than one insecticide [@GPIRM2012] but until there are more data on whether behavioural resistance is genetic and heritable we would not recommend representing it in this way.
